Taking uncertainty in word embedding bias estimation seriously: a bayesian approach

### 2021-12-10 paperDraft/paperDraft2.tex
setting up paper draft 2

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  10pt,
  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Taking uncertainty in word embedding bias estimation seriously: a bayesian approach},
  pdfauthor={Alicja Dobrzeniecka and Rafal Urbaniak},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
%\documentclass{article}

% %packages
 \usepackage{booktabs}

\usepackage{multirow}

@@

\usepackage[textsize=footnotesize]{todonotes}
%\linespread{1.5}

@@

%language
\usepackage{times}
\usepackage{t1enc}
%\usepackage[utf8x]{inputenc}
%\usepackage[polish]{babel}
%\usepackage{polski}
\usepackage[utf8]{inputenc}
\usepackage{mathptmx}
\usepackage[scaled=0.88]{helvet}

@@

%abbreviations
\newcommand{\ra}{\rangle}
\newcommand{\la}{\langle}
\newcommand{\n}{\neg}
\newcommand{\et}{\wedge}
\newcommand{\jt}{\rightarrow}
\newcommand{\ko}[1]{\forall  #1\,}
\newcommand{\ro}{\leftrightarrow}
\newcommand{\exi}[1]{\exists\, {_{#1}}}
\newcommand{\pr}[1]{\mathsf{P}(#1)}
\newcommand{\cost}{\mathsf{cost}}

@@

\newtheorem{q}{\color{blue}Question}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

@@

	\rule{10cm}{0.4pt}\end{center}
	\end{minipage}
	}
%--------------------

@@

%\bibliographystyle{apalike}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Taking uncertainty in word embedding bias estimation seriously: a
bayesian approach}
\author{Alicja Dobrzeniecka and Rafal Urbaniak}
\date{}

@@

@@

One of the first measures in the discussion has been developed by
(\textbf{bolukbasi2016man?}). There, the gender bias of a word \(w\) is
understood as its projection on the gender direction
\(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the
gender direction is the top principal compontent of ten gender pair
difference vectors). The underlying idea is that no bias is present if
non-explicitly gendered words are in equal distance to both elements in
all explicitly gender pairs. Given the (ideally) gender neutral words
\(N\) and the gender direction \(g\) the direct gender bias is defined
as the average distance of the words in \(N\) from \(g\) (\(c\) is a
parameter determining how strict we want to be): \begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

The use of projections has been ciriticized for instance by
(\textbf{Gonen2019lipstick?}), who point out that while gender-direction
might be an indicator of bias, it is only one possible manifestation of
it, and reducing a projection of words might be insufficient. For
instance, ``math'' and ``delicate'' might be in equal distance to both
explicitly gendered words while being closer to quite different
stereotypical attribute words. Further, the authors point out that most
word pairs preserve similarity under debiasing meant to minimize
projection-based bias.\footnote{(\textbf{bolukbasi2016man?}) use also
  another method which involves analogies and their evaluations by human
  users on Mechanical Turk. It is discussed and criticized in
  (\textbf{Nissim2020fair?}).}

To measure bias in word embeddings,
(\textbf{Caliskan2017semanticsBiases?}) proposed the Word Embedding
Association Test (WEAT). The idea is that the measure of biases between
two sets of target words, \(X\) and \(Y\), (we call them protected
words) should be quantified in terms of the cosine similarity between
the protected words and attribute words coming from two sets of
stereotype attribute words, \(A\) and \(B\) (we'll call them
attributes). For instance, \(X\) might be a set of male names, \(Y\) a
set of female names, \(A\) might contain stereotypically male-related
career words, and \(B\) stereotypically female-related family words.
WEAT is a modification of the Implicit Association Test (IAT)
(\textbf{Nosek2002harvesting?}) used in psychology and uses almost the
same word sets, allowing for a \emph{prima facie} sensible comparison
with bias in humans. If the person's attitude towards given pair of
concept is to be interpreted as neutral, there should be no noticeable
task completion time difference, and the final value from the formula
should be around 0. Let \(f\) be a similarity measure (usually, cosine
similarity). The association difference for a term \(t\) is:
\begin{align}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\end{align} \noindent then, the association difference between \(A\) a
\(B\) is: \begin{align}
s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
\end{align} \noindent \(s(X,Y,A,B)\) is the statistic used in the
signifcance test, and the \(p\)-value obtained by bootstrapping: it is
the frequency of \(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized
partitions \(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by
normalizing the difference in means as follows: \begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

(\textbf{Caliskan2017semanticsBiases?}) show that significant
biases---thus measured--- similar to the ones discovered by IAT can be
discovered in word embeddings. (\textbf{Lauscher2019multidimensional?})
extended the methodology to a multilingual and cross-lingual setting,
arguing that using Euclidean distance instead of similarity does not
make much difference, while the bias effects vary greatly across
embedding models (interestingly, with social media-text trained
embeddings being less biased than those based on Wikipedia).

A similar methodology is employed by (\textbf{Garg2018years?}), who
employ word embeddings trained on corpora from different decades to
study the shifts in various biases. For instance, to compute the
occupational embeddings bias for women the authors first compute the
average vector of vector emeddings of words that represent women
(e.g.~``she,'' ``female''), then calculate the Euclidean distance
between this mean vector and words for occupations. Then they take the
mean of these distances and subtract from it the analogously obtained
mean for the average vector of vector embeddings of words that represent
men. Formally they take the relative norm distance between \(X\) and
\(Y\) to be: \begin{align}
\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
\end{align} \noindent where the norm used is Euclidean, and \(v_X\) and
\(x_Y\) are average vectors for sets \(X\) and \(Y\) respectively.

@@

\hypertarget{some-methodological-problems}{%
\section{Some methodological
problems}\label{some-methodological-problems}}

Such approaches, however, are not unproblematic. One may cosinder the
selection of attributes for the bias research. For one thing, the word
lists are often fairly small, and the papers do not discuss the impact
of word list sizes on the uncertainty involved. We will argue that the
word list sizes are too small to yield strong conclusions, given the
variances involved.

Crucially, these approaches use means of mean average cosine
similarities to measure similarity between protected word and harmful
stereotypes. If one takes a closer look at the individual values that
are taken for the calculations, it turns out that there are quite a few
outliers and surprisingly dissimilar words. This issue will become
transparent when we inspect the visualizations of individual cosine
distances, following the idea that one of the first step to understand
data is to look at it.

With such a method the uncertainty involved is not really considered
which makes it even more difficult to give reasonable interpretations of
the results. We propose the use of Bayesian method to obtain some
understanding of the influence the uncertainty has on the interpretation
of final results.

We also observed that in the paper there is no word class distinction
and no control group. In the original paper, words from all three
religions were compared against all of the stereotypes, which means that
there was no distinction between cases in which the stereotype is
associated with a given religion, as opposed to the situation in which
it is associated with another one. Not all of the stereotypical words
have to be considered as harmful for all of the religions. One should
investigate the religions separately as some of them may have stronger
harmful associations that others. One should also include control groups
to have a way of comparing the stereotypical results with neutral or
human-like words. Later in the text we will explain in details reasons
for introducing control groups. In our analysis, we distinguish between
stereotypes associated with a given group, stereotypes associated with
different groups, and control groups: neutral words and stereotypes-free
human predicates.

The interpretation of the results is also a challenge. Assuming for a
moment that the value of multi-class cosine distance is correct, one may
question the interpretation. Manzini, Lim, Tsvetkov, \& Black (2019)
summarize the averages of cosine distance per group (gender, race,
religion). For now let us focus now on analyzing the values relating to
religious biases. Here is the relevant fragment of table:

\begin{longtable}[]{@{}ll@{}}
\toprule
Religion Debiasing & MAC \\
\midrule
\endhead
Biased & 0.859 \\
Hard Debiased & 0.934 \\
Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
\bottomrule
\end{longtable}

MAC stand for mean average cosine similarity, although in reality the
the table contains mean distances. What may attract attention is the
fact that the value of cosine distance in ``Biased'' category is already
quite high even before debiasing. High cosine distance indicates low
cosine similarity between values. One could think that average cosine
similarity equal to approximately 0.141 is not significant enough to
consider it as bias. However the authors aim to mitigate ``biases'' in
vectors with such great distance to make it even larger.
Methodologically the question is, on what basis is this small similarity
still considered as a proof of the presence of bias, and whether these
small changes are meaningful. This is in general the problem of scale
and the lack of universal intervals. In contrast, statistical intervals
will help us decide whether a given cosine similarity is high enough to
consider the words to be more similar than if we chose them at random.
We will use highest posterior density intervals, in line with Bayesian
methodology.

One may consider the issue of the curse of dimensionality. In our case,
the curse of dimensionality may take place when there is an increase in
the volume of data that results in adding extra dimensions to the
Euclidean space. As the number of features increases, it may be harder
and harder to obtain useful information from the data using the
available algorithms. Using cosine similarity in high dimensions in word
embeddings may be prone to the curse of dimensionality. According to
Venkat (2018) there are reasons to consider this phenomenon when
searching for word similarities in higher dimensions. An experiment is
conducted that aims at showing how the similarity values and variation
change as the number of dimensions increases. The hypothesis made in the
paper states that two things will happen as the number of dimensions
increase. First, the effort required to measure cosine similarity will
be greater, and two, the similarity between data will blur out and have
less variation. The authors generate random points with increasing
number of dimensions where each dimension of a data point is given a
value between 0 and 1. Then they pick one vector at random from each
dimension class and calculate the cosine similarity between the chosen
vector and the rest of the data. Then they check how the variation of
values changes as the number of dimensions increases. It seems like the
more dimensions there are, the smaller the variance and therefore it is
less obvious how to interpret the resulting cosine similarities. Maybe
the scale should be adjusted to the number of dimensions and variance so
that it still gives us sensible information about data.

\hypertarget{the-problem-with-means-of-means}{%
\section{The problem with means of
means}\label{the-problem-with-means-of-means}}

The measures described all calculate means of means and their authors
run statistical tests on sets of means. This, however, is problematic
for two related reasons. One, by pre-averaging data we throw away
information about sample sizes. For the former point, think about
proportions: 10 out of 20 and 2 out of 4 give the same mean, but you
would obtain more information by making the former observation rather
than by making the latter. Two, when we pre-average, we remove
variation, and so pre-averaging tend to manufacture false confidence. We
will have more to say about whether this happens in the case of
applications of MAC, for now let's go over a simple example to make the
conceptual point clear.

To illustrate let's employ the formulas used by
(\textbf{Caliskan2017semanticsBiases?}) in a simple example.
Conceptually, all such tests come up with rather short lists of
protected words and rather short lists of stereotypical attributes.
Clearly, these are not complete list. So let's treat them as samples
from richer pools of stereotypical predicates and let's take the
uncertainty and variation involved seriously.

Consider a simple situation in which there are two protected classes,
\(X=\{t_1,t_2\}\) and \(Y=\{t_3,t_4\}\) and two five-element attribute
sets \(A\) and \(B\).

First, we play around with a scenario in which all the protected terms
are on average equally unsimilar to both sets (\(\mu =0\)) with standard
deviation of \(.05\). Let's randomly draw similarity scores and plot the
results with group means plotted as vertical lines.

\footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
\NormalTok{t1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
\NormalTok{t2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
\NormalTok{t3 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
\NormalTok{t4 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\normalsize

\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-2-1} \end{center}

\noindent When you look at the datapoints, do you have the impression of
a strong bias here? We wouldn't think so. But now let's run the
calculations from (\textbf{Caliskan2017semanticsBiases?}).

\vspace{1mm}
\footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{s }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{ (table)\{ }\FunctionTok{mean}\NormalTok{(table}\SpecialCharTok{$}\NormalTok{A) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(table}\SpecialCharTok{$}\NormalTok{B)\}}
\NormalTok{factor }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2),}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4)))}
\NormalTok{numerator }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4))}
\FunctionTok{print}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{factor =}\NormalTok{ factor,}\AttributeTok{numerator =}\NormalTok{ numerator, }\AttributeTok{bias =}\NormalTok{ numerator }\SpecialCharTok{/}\NormalTok{ factor))}
\end{Highlighting}
\end{Shaded}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  10pt,
  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\title{Taking uncertainty in word embedding bias estimation seriously: a
bayesian approach}
\author{Alicja Dobrzeniecka and Rafal Urbaniak}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Taking uncertainty in word embedding bias estimation seriously: a bayesian approach},
  pdfauthor={Alicja Dobrzeniecka and Rafal Urbaniak},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
%\documentclass{article}

% %packages
\usepackage{booktabs}
%\usepackage[left]{showlabels}
\usepackage{multirow}
\usepackage{subcaption}

@@

\usepackage[textsize=footnotesize]{todonotes}
\newcommand{\ali}[1]{\todo[color=gray!40]{\textbf{Alicja:} #1}}
\newcommand{\mar}[1]{\todo[color=blue!40]{#1}}
\newcommand{\raf}[1]{\todo[color=olive!40]{#1}}

%\linespread{1.5}
\newcommand{\indep}{\!\perp \!\!\! \perp\!}

@@

%language
%\usepackage{times}
\usepackage{mathptmx}
\usepackage[scaled=0.86]{helvet}
\usepackage{t1enc}
%\usepackage[utf8x]{inputenc}
%\usepackage[polish]{babel}
%\usepackage{polski}

@@

%abbreviations
\newcommand{\ra}{\rangle}
\newcommand{\la}{\langle}
\newcommand{\n}{\neg}
\newcommand{\et}{\wedge}
\newcommand{\jt}{\rightarrow}
\newcommand{\ko}[1]{\forall  #1\,}
\newcommand{\ro}{\leftrightarrow}
\newcommand{\exi}[1]{\exists\, {_{#1}}}
\newcommand{\pr}[1]{\ensuremath{\mathsf{P}(#1)}}
\newcommand{\cost}{\mathsf{cost}}
\newcommand{\benefit}{\mathsf{benefit}}
\newcommand{\ut}{\mathsf{ut}}

@@

\newtheorem{q}{\color{blue}Question}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}[fact]

@@

	\rule{10cm}{0.4pt}\end{center}
	\end{minipage}
	}

%--------------------

@@

%\bibliographystyle{apalike}

\usepackage{tikz}
\usetikzlibrary{positioning,shapes,arrows}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\raf{fix bibliography}

@@

One of the first measures in the discussion has been developed by
Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016). There, the gender
bias of a word \(w\) is understood as its projection on the gender
direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
(the gender direction is the top principal compontent of ten gender pair
difference vectors). The underlying idea is that no bias is present if
non-explicitly gendered words are in equal distance to both elements in
all explicitly gender pairs. Given the (ideally) gender neutral words
\(N\) and the gender direction \(g\) the direct gender bias is defined
as the average distance of the words in \(N\) from \(g\) (\(c\) is a
parameter determining how strict we want to be): \begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

The use of projections has been ciriticized for instance by Gonen \&
Goldberg (2019), who point out that while gender-direction might be an
indicator of bias, it is only one possible manifestation of it, and
reducing a projection of words might be insufficient. For instance,
``math'' and ``delicate'' might be in equal distance to both explicitly
gendered words while being closer to quite different stereotypical
attribute words. Further, the authors point out that most word pairs
preserve similarity under debiasing meant to minimize projection-based
bias.\footnote{Bolukbasi et al. (2016) use also another method which
  involves analogies and their evaluations by human users on Mechanical
  Turk. It is discussed and criticized in (Nissim, Noord, \& Goot,
  2020).}

To measure bias in word embeddings, Caliskan, Bryson, \& Narayanan
(2017) proposed the Word Embedding Association Test (WEAT). The idea is
that the measure of biases between two sets of target words, \(X\) and
\(Y\), (we call them protected words) should be quantified in terms of
the cosine similarity between the protected words and attribute words
coming from two sets of stereotype attribute words, \(A\) and \(B\)
(we'll call them attributes). For instance, \(X\) might be a set of male
names, \(Y\) a set of female names, \(A\) might contain stereotypically
male-related career words, and \(B\) stereotypically female-related
family words. WEAT is a modification of the Implicit Association Test
(IAT) (Nosek, Banaji, \& Greenwald, 2002) used in psychology and uses
almost the same word sets, allowing for a \emph{prima facie} sensible
comparison with bias in humans. If the person's attitude towards given
pair of concept is to be interpreted as neutral, there should be no
noticeable task completion time difference, and the final value from the
formula should be around 0. Let \(f\) be a similarity measure (usually,
cosine similarity). The association difference for a term \(t\) is:
\begin{align}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\end{align} \noindent then, the association difference between \(A\) a
\(B\) is: \begin{align}
s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
\end{align} \noindent \(s(X,Y,A,B)\) is the statistic used in the
significance test, and the \(p\)-value obtained by bootstrapping: it is
the frequency of \(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized
partitions \(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by
normalizing the difference in means as follows: \begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

Caliskan et al. (2017) show that significant biases---thus measured---
similar to the ones discovered by IAT can be discovered in word
embeddings. Lauscher \& Glavas (2019) extended the methodology to a
multilingual and cross-lingual setting, arguing that using Euclidean
distance instead of similarity does not make much difference, while the
bias effects vary greatly across embedding models (interestingly, with
social media-text trained embeddings being less biased than those based
on Wikipedia).

A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
(2018), who employ word embeddings trained on corpora from different
decades to study the shifts in various biases. For instance, to compute
the occupational embeddings bias for women the authors first compute the
average vector of vector embeddings of words that represent women
(e.g.~``she,'' ``female''), then calculate the Euclidean distance
between this mean vector and words for occupations. Then they take the
mean of these distances and subtract from it the analogously obtained
mean for the average vector of vector embeddings of words that represent
men. Formally they take the relative norm distance between \(X\) and
\(Y\) to be: \begin{align}
\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
\end{align} \noindent where the norm used is Euclidean, and \(v_X\) and
\(x_Y\) are average vectors for sets \(X\) and \(Y\) respectively.

@@
